---
title: Time Series Analysis Methods and Applications
author: 
  - name          : "Igor Baranov"
    affiliation   : "York University School of Continuing Studies"
    email         : "https://learn.continue.yorku.ca/user/profile.php?id=21219"
  - name          : "Michael Parravani"
    affiliation   : "York University School of Continuing Studies"
  - name          : "Ariana Biagi"
    affiliation   : "York University School of Continuing Studies"
  - name          : "Hui Fang Cai"
    affiliation   : "York University School of Continuing Studies"
abstract: >
  The goal of this project is to discover Time Seties analysis modelsand algorithms, show different application.
output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex
figsintext        : no
---

# Introduction

## Background
## Objective
## Plan
## Ethical ML Framework
As the goal of this application is only to recommend jokes, many aspects of the ethical ML framework do not directly apply. The data is open source and we can assume was collected in transparent ways. That being said, there is likely a large segment of the populace that is under represented in this ratings dataset - we assume a low income population (limited access to internet, limited time to be spent rating jokes, etc). This will potentially reduce the reccomender's accuracy for that group of the population. If the outcome of this system were to be of more social impact, this would need to be corrected with appropriate data collection methods.

# Data understanding

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
set.seed(777)
library("ggplot2")
library("readr")
```

```{r eval=FALSE, include=FALSE}
# Set directory to current script (works in R Studio only)
this.dir <- dirname(rstudioapi::getActiveDocumentContext()$path)
setwd(this.dir)
```

## Dataset 1
Yahoo Science labeled time series. This set is big, it should be downloaded and used locally:
[Yahoo Science labeled time series](https://github.com/ivbsoftware/CSDA1040-project-3/blob/master/data/orig/ydata-labeled-time-series-anomalies-v1_0.zip)

## Dataset 2
NAB Data Corpus: better just load from github directly to R script. [NAB Data Corpus](https://github.com/numenta/NAB/tree/master/data)

## Data preparation

## Load the jokes rating data and do the basic checks
```{r message=FALSE, warning=FALSE}
df <- read_csv(file="../../../data/jesterfinal151cols.csv.zip",col_names = FALSE)
dim(df)
```

# Prepare Shiny App for Deployment

## Saving recommender data objects
```{r message=FALSE, warning=FALSE}
saveRDS(rcmnd_ub, file = "jokeRecommender.Rds")
saveRDS(jokes, file = "jokes.Rds")
```

# Deployment Discussion

This model is currently not of much use given its accuracy but it will serve as a proof of concept. This model could be used to help writers of movies/tv shows write jokes appropriate for a specific or large audience. 

More data should be collected from this userbase to fill a training dataset. The dataset in its current state is quite sparse. The data would need to be updated every 3-5 years as people's taste changes and people within certian age groups mature. The Shiny app developed would be a deployment method to collect more data. 

Further analysis could be done (with the appropriate data) to see how similar taste in humor is related to age.

The model developed in this project was used to create Shiny application currently deployed at [ivbsoftware.shinyapps.io/JokeRecommender/](https://ivbsoftware.shinyapps.io/TimeSeries1/). Code of the application could be found in [Github](https://github.com/ivbsoftware/CSDA1040-project-3/tree/master/shiny/).

\bibliography{RJreferences}

\newpage

# Note from the Authors

This file was generated using [_The R Journal_ style article template](https://github.com/rstudio/rticles), additional information on how to prepare articles for submission is here - [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf). The article itself is an executable R Markdown file that could be [downloaded from Github](https://github.com/ivbsoftware/CSDA1040-project-1/tree/master/scripts/R_Journal/csda1040-lab1) with all the necessary artifacts.