---
title: "Financial Fraud Detection"
author: "Igor Baranov"
date: "February 13, 2019"
output: pdf_document
---

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)

library("ggplot2")
library("readr")
library("dplyr")
library("forcats")

set.seed(777)
```

```{r eval=FALSE, include=FALSE}
# Seting directory to current script (works in R Studio only)
#this.dir <- dirname(rstudioapi::getActiveDocumentContext()$path)
#setwd(this.dir)
```

# Introduction
Paysim synthetic dataset is borrowed from Kaggle's [Synthetic Financial Datasets For Fraud Detection](https://www.kaggle.com/ntnu-testimon/paysim1)of mobile money transactions. Each step represents an hour of simulation. This dataset is scaled down 1/4 of the original dataset which is presented in the paper "PaySim: A financial mobile money simulator for fraud detection".

# Preparing data
## Load and clean 
The data should be downloaded from Kaggle's [Synthetic Financial Datasets For Fraud Detection](https://www.kaggle.com/ntnu-testimon/paysim1/downloads/PS_20174392719_1491204439457_log.csv/2).

Loading and converting some colimns to factors:

```{r message=FALSE, warning=FALSE}
if(!file.exists("../data/financial-fraud-detection-ib/dff.Rds")) {
  dff <- read_csv(file="../data/orig/PS_20174392719_1491204439457_log.csv.zip",col_names = TRUE)
  
  dff <- dplyr::tbl_df(dff)
  dplyr::glimpse(dff)
  utils::View(dff)
  
  
  dff$type <- as.factor(dff$type)
  dff$nameOrig <- as.factor(dff$nameOrig)
  dff$nameDest <- as.factor(dff$nameDest)
  saveRDS(dff, file = "../data/financial-fraud-detection-ib/dff.Rds")
  gc()
} else {
  dff <- readRDS("../data/financial-fraud-detection-ib/dff.Rds")
}
head(dff)

# Feature engineering

## Distribution of number of tranactions by 'nameOrig'

```{r fig.height=7, fig.width=5.5, message=FALSE, warning=FALSE}
nameOrig_freq_sorted_list <- levels(fct_infreq(dff$nameOrig))


ggplot() + aes(dff$nameOrig)+ geom_histogram(binwidth=10, colour="black", fill="white", stat = "count")

ggplot() + aes(fct_infreq(dff[dff$nameOrig %in% nameOrig_freq_sorted_list[1:30],]$nameOrig))+ 
  geom_histogram(colour="black", fill="white", stat = "count")+
  ylab("Frequency") + xlab("")+
  theme(axis.text.x = element_text(angle =90, hjust = 1))
``````